---
title: "ModelR: a workflow for ecological niche models based on dismo"
author: "Andrea Sánchez-Tapia"
date: "`r Sys.Date()`"
output: 
    rmarkdown::html_vignette:
        toc: true
vignette: >
  %\VignetteIndexEntry{ModelR: a workflow for ecological niche models based on dismo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


ModelR is a workflow based on dismo designed to automatize some of the common steps when performing ecological niche models. Given the occurrence records and a set of environmental predictors, it prepares the data to crossvalidation or jacknife procedures depending on the number of occurrence points<!-- precisamos implementar bootstrap-->, then it performs ecological niche models using the several algorithms implemented in the `dismo` package.

The workflow consists of mainly three functions that should be used sequentially.

1. `do_enm()` and each independent `do_[algorithm]` function make the ENM for each partition of the crossvalidation
2. `final_model()` selects and joins the partition models into a final model per species per algorithm, 
3. `ensemble_model()` joins the final models per algorithm into an ensemble model.

## Folder structure

`modelr` writes the outputs in the hard disk, according to the following folder structure:   

    `models.dir/projection/partitions`  
    `models.dir/projection/final_models`  
    `models.dir/projection/ensemble_models`  

+ When projecting models into the present, the projection folder is called `present`.
+ You can set `models.dir` wherever you want in the hard disk, but if you do not modify the default value, it will create the output under the working directory (its default value is `./models.dir`, the period points to the working directory)
+ The nested subfolder structure will remain the same. If you change `final_models` default value ("final_model") you will need to include the new value when calling ensemble() (`final_dir = "[new name]"`), to indicate the function where to look for models.


# Fitting a model per partition

All `do_[algorithm]` functions and `do_enm()` create a *model per partition, per algorithm*. The available algorithms are bioclim, domain, maxent, mahalanobis distances, as implemented in `dismo`, and support vector machines (SVM), as implemented by packages `kernlab` and `e1071`. GLM and Random Forest (from package `randomForest`) are also implemented.

```{r load}
devtools::load_all()
```

Other options in do_enm and individual algorithm functions are: 

+ `partitions`: It implements a k-fold cross-validation (argument `part`, defaults to 3) but overwrites part when n < 10, setting part to the number of occurrence records (a jacknife partition).  
+ `buffer`
+ seed 
+ mask








```{r load_java, echo = FALSE}
dyn.load(
    '/Library/Java/JavaVirtualMachines/jdk1.8.0_144.jdk/Contents/Home/jre/lib/server/libjvm.dylib')
```

```{r especies}
library(rJava) 
library(raster)
especies <- unique(coordenadas$sp)
especies
```

## Fitting one algorithm: `do_bioclim`

### Setting up the data

`modelr` comes with example data, a data frame called coordenadas, with occurrence data for four species, and predictor variables called `variaveis_preditoras`

```{r dataset, fig.width= 5, fig.height=5, fig.cap= "The example dataset: predictor variables and occurrence for four species "}
head(coordenadas)
raster::plot(variaveis_preditoras[[1]])
points(sp::SpatialPoints(coordenadas[,c(2,3)]), 
       bg = as.numeric(unclass(coordenadas$sp)), pch = 21)
```


We'll filter the coordenadas file to select only the data for the first species: 

```{r occs}
library(dplyr)
especies[1]
occs <- filter(coordenadas, sp == especies[1]) %>% select(lon, lat)
head(occs)
```

```{r do_bioclim, cache = T}
args(do_bioclim)
do_bioclim(sp = especies[1],
           coordinates = occs,
           partitions = 3,
           buffer = "mean",
           predictors = variaveis_preditoras,
           mask = NULL,
           models.dir = "~/temp",
           write_png = T,
           n.back = 500)
```


```{r files}
list.files("~/temp", recursive = T)
```
 
At the end of a model round, the partition folder containts: 

+ A file called `sdmdata.txt` with the data used for each partition. 
+ Figures in png to explore the results readily, without reloading them into R or opening them in a SIG program. The creation of these .png can be controlled with the `write_png` parameter. 
+ A tif file for each partition
+ A .txt table with the evaluation data for each partition: `evaluate_[Species name ]_[partition number]_[algorithm].txt`. These files will be read by the `final_model()` function, to generate the final model per species.


### Fitting several algorithms per species: do_enm()

The same modeling procedure can be performed by using do_enm(), that recevives the same parameters as do_bioclim() but allows the user to call the algorithms using TRUE or FALSE statements (just as BIOMOD2 functions do). 

```{r, eval = F}
args(do_enm)
do_enm(sp = especies[1],
       coordinates = coordenadas,
       partitions = 3,
       buffer = "mean",
       predictors = variaveis_preditoras,
       mask = NULL,
       models.dir = "~/temp",
       write_png = T,
       n.back = 500,
       bioclim = T)
```

`do_enm()` calls several independent functions for each algorithm, called `do_[algorithm]`. This allows to parallelize by species and/or species algorithms in multi-cluster environments. The following lines call for bioclim, GLM, maxent, random forests and smv.

```{r, eval = T}
args(do_enm)
do_enm(sp = especies[1],
       coordinates = occs,
       partitions = 3,
       buffer = "mean",
       predictors = variaveis_preditoras,
       mask = NULL,
       models.dir = "~/temp",
       write_png = T,
       n.back = 500,
       bioclim = T, 
       glm = T,
       maxent = T,
       rf = T,
       svm = T)
```


## Joining partitions: `final_model()`

`final_model()` has two main ways to create a final model per algorithm per species.

+ It can select the best partitions, with the parameter `select.partitions = T`, and select only those who obtained a TSS value above `TSS.value`. If select.partitions is set to FALSE, it will create the final model using all partitions. 
+ It can also weigh the partitions by a performance metric, giving larger weights to better partitions.
+ The final models can be done using a subset of the algorithms avaliable on the kard disk, using the parameter `algorithms`. If left unspecified, all algorithms will be used.
+ 



```{r final_model}
args(final_model)
final_model(sp = especies[1],
            select.partitions = F,
            TSS.value = 0.5,
            algoritmos = c("BioClim", "glm", "maxent"),
            models.dir = "~/temp")
```

`final_model()` creates a .tif fil for each final.model
 
 ```{r}
final.folder <- list.files("~/temp", recursive = T, pattern = "final_models", include.dirs = T, full.names = T)
list.files(final.folder)
```

## ensemble()

```{r}
ensemble(especies[1],
         occs = occs,
         models.dir = "~/temp",
         consensus = T,
         consensus.level = 0.5)
```


